### 2021-03-28 final_report/include/TT.tex
checked bibliography entries + minor changes
capitalized footnotes

While many artificial languages, such as programming languages, are compositional by construction, the general validity of this principle for natural languages is, despite many arguments in favor, under debate. 
In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{in Section \ref{evalign}, however, we will see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}. \smallskip

@@

% cons
On the other hand, there is currently no effective way to use these trees as a starting point for NLG: dependency grammars are an effective way to describe language, but, unlike phrase-structure grammars, they are not \textit{generative},\footnote{that does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}. \smallskip

@@

In the following, we give a quick overview of the standard format UD uses to store dependency trees and of the aspects of the annotation scheme that are most relevant to the task at hand. 
Appendix \ref{a} provides a more comprehensive, but not completely exhaustive, description of all the POS tags and dependency labels mentioned in the examples appearing in this work based on that in \cite{compsyn}. 
The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{available at \url{universaldependencies.org}.}.

\paragraph{The CoNLL-U format} \label{conll}
The standard plain text format for dependency trees, CoNLL-U, is an extension of CoNNL-X \cite{conllx} which may contain \textit{comment lines}, starting with \texttt{\#}, \textit{blank lines} marking sentence boundaries and \textit{word lines} containing the annotation of a single token (generally a word, with some exceptions\footnote{examples of such exceptions are Italian contractions such as ``\textit{dello}'', which is generally divided into ``\textit{del}'' + ``\textit{lo}''.} in 10 tab-separated fields: \smallskip

\begin{enumerate}
    \item \texttt{ID}: integer\footnote{again, there are exceptions: in the case of Italian contractions and other multiword tokens, ranges may be used. Furthermore, empty nodes are characterized by decimal numbers between 0 and 1.} representing the position of the word in the sentence, starting from 1
    \item \texttt{FORM}: inflected from of the word
    \item \texttt{LEMMA}: lemmatized form of the word
    \item \texttt{UPOS}: universal POS tag
    \item \texttt{XPOS}: language-specific POS tag (optional)
    \item \texttt{FEATS}: list of (universal or language-specific) morphological features (optional)
    \item \texttt{HEAD}: head of the current word, i.e. either the value of the \texttt{ID} of another word in the same sentence or 0 in case the word at hand is a \texttt{root}
    \item \texttt{DEPREL}: universal dependency relation to the \texttt{HEAD} (\texttt{root} in case the word at hand is itself the \texttt{root})
    \item \texttt{DEPS}: enhanced dependency graph in the form of a list of \texttt{HEAD}-\texttt{DEPREL} pairs (optional)
    \item \texttt{MISC}: any other annotation (optional)
\end{enumerate} \smallskip

\begin{figure}[h]
    \centering
    \begin{subfigure}{.55\textwidth}
      \centering
      \footnotesize
        \begin{verbatim}
1	this	this	PRON	_	_	5	nsubj	_	_
2	is	be	AUX	_	_	5	cop	_	_
3	a	a	DET	_	_	5	det	_	_
4	dependency	dependency	NOUN	_	_	5	compound	_	_
5	tree	tree	NOUN	_	_	0	root	_	_
        \end{verbatim}
    \end{subfigure}%
    \begin{subfigure}{.45\textwidth}
        \centering
        \footnotesize
        \setlength{\unitlength}{0.21mm}
        \begin{picture}(371.0,130.0)
            \put(0.0,0.0){this}
            \put(46.0,0.0){is}
            \put(83.0,0.0){a}
            \put(120.0,0.0){dependency}
            \put(220.0,0.0){tree}
            \put(0.0,15.0){{\tiny PRON}}
            \put(46.0,15.0){{\tiny AUX}}
            \put(83.0,15.0){{\tiny DET}}
            \put(120.0,15.0){{\tiny NOUN}}
            \put(220.0,15.0){{\tiny NOUN}}
            \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
            \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
            \put(105.0,99.66666666666667){{\tiny nsubj}}
            \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
            \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
            \put(128.0,83.0){{\tiny cop}}
            \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
            \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
            \put(146.5,66.33333333333334){{\tiny det}}
            \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
            \put(131.5,35.0){\vector(0,-1){5.0}}
            \put(165.0,49.66666666666667){{\tiny compound}}
            \put(235.0,130.0){\vector(0,-1){100.0}}
            \put(240.0,120.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A dependency tree in CoNNL-U format alongside its graphical representation]{A dependency tree in CoNNL-U format alongside its graphical representation. Optional fields in the CoNLL-U file are left blank.}
    \label{deptree}
\end{figure}

@@

An important distinction we can make based on POS tags is, as we will see in Section \ref{criteria}, that between \textit{content} and \textit{function} words.
\textit{Content} or \textit{open class} words are words with a lexical meaning (nouns, lexical verbs, adjectives, adverbs and interjections).
\textit{Function} words, like pronouns and determiners, only have a grammatical meaning. Since they do not readily accept new members\footnote{one exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}, they are often referred to also as \textit{closed class} words. \smallskip

>>>>

While many artificial languages, such as programming languages, are compositional by construction, the general validity of this principle for natural languages is, despite many arguments in favor, under debate. 
In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{In Section \ref{evalign}, however, we will see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}. \smallskip

@@

% cons
On the other hand, there is currently no effective way to use these trees as a starting point for NLG: dependency grammars are an effective way to describe language, but, unlike phrase-structure grammars, they are not \textit{generative},\footnote{That does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}. \smallskip

@@

In the following, we give a quick overview of the standard format UD uses to store dependency trees and of the aspects of the annotation scheme that are most relevant to the task at hand. 
Appendix \ref{a} provides a more comprehensive, but not completely exhaustive, description of all the POS tags and dependency labels mentioned in the examples appearing in this work based on that in \cite{compsyn}. 
The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{Available at \url{universaldependencies.org}.}.

\paragraph{The CoNLL-U format} \label{conll}
The standard plain text format for dependency trees, CoNLL-U, is an extension of CoNNL-X \cite{conllx} which may contain \textit{comment lines}, starting with \texttt{\#}, \textit{blank lines} marking sentence boundaries and \textit{word lines} containing the annotation of a single token (generally a word, with some exceptions\footnote{Examples of such exceptions are Italian contractions such as ``\textit{dello}'', which is generally divided into ``\textit{del}'' + ``\textit{lo}''.} in 10 tab-separated fields: \smallskip

\begin{enumerate}
    \item \texttt{ID}: integer\footnote{Again, there are exceptions: in the case of Italian contractions and other multiword tokens, ranges may be used. Furthermore, empty nodes are characterized by decimal numbers between 0 and 1.} representing the position of the word in the sentence, starting from 1
    \item \texttt{FORM}: inflected from of the word
    \item \texttt{LEMMA}: lemmatized form of the word
    \item \texttt{UPOS}: universal POS tag
    \item \texttt{XPOS}: language-specific POS tag (optional)
    \item \texttt{FEATS}: list of (universal or language-specific) morphological features (optional)
    \item \texttt{HEAD}: head of the current word, i.e. either the value of the \texttt{ID} of another word in the same sentence or 0 in case the word at hand is a \texttt{root}
    \item \texttt{DEPREL}: universal dependency relation to the \texttt{HEAD} (\texttt{root} in case the word at hand is itself the \texttt{root})
    \item \texttt{DEPS}: enhanced dependency graph in the form of a list of \texttt{HEAD}-\texttt{DEPREL} pairs (optional)
    \item \texttt{MISC}: any other annotation (optional)
\end{enumerate} \smallskip

\begin{figure}[h]
    \centering
    \begin{subfigure}{.55\textwidth}
      \centering
      \footnotesize
        \begin{verbatim}
1	this	this	PRON	_	_	5	nsubj	_	_
2	is	be	AUX	_	_	5	cop	_	_
3	a	a	DET	_	_	5	det	_	_
4	dependency	dependency	NOUN	_	_	5	compound	_	_
5	tree	tree	NOUN	_	_	0	root	_	_
        \end{verbatim}
    \end{subfigure}%
    \begin{subfigure}{.45\textwidth}
        \centering
        \footnotesize
        \setlength{\unitlength}{0.21mm}
        \begin{picture}(371.0,130.0)
            \put(0.0,0.0){this}
            \put(46.0,0.0){is}
            \put(83.0,0.0){a}
            \put(120.0,0.0){dependency}
            \put(220.0,0.0){tree}
            \put(0.0,15.0){{\tiny PRON}}
            \put(46.0,15.0){{\tiny AUX}}
            \put(83.0,15.0){{\tiny DET}}
            \put(120.0,15.0){{\tiny NOUN}}
            \put(220.0,15.0){{\tiny NOUN}}
            \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
            \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
            \put(105.0,99.66666666666667){{\tiny nsubj}}
            \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
            \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
            \put(128.0,83.0){{\tiny cop}}
            \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
            \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
            \put(146.5,66.33333333333334){{\tiny det}}
            \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
            \put(131.5,35.0){\vector(0,-1){5.0}}
            \put(165.0,49.66666666666667){{\tiny compound}}
            \put(235.0,130.0){\vector(0,-1){100.0}}
            \put(240.0,120.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A dependency tree in CoNNL-U format alongside its graphical representation]{A dependency tree in CoNNL-U format alongside its graphical representation. Optional fields in the CoNLL-U file are left blank.}
\end{figure}

@@

An important distinction we can make based on POS tags is, as we will see in Section \ref{criteria}, that between \textit{content} and \textit{function} words.
\textit{Content} or \textit{open class} words are words with a lexical meaning (nouns, lexical verbs, adjectives, adverbs and interjections).
\textit{Function} words, like pronouns and determiners, only have a grammatical meaning. Since they do not readily accept new members\footnote{One exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}, they are often referred to also as \textit{closed class} words. \smallskip
